---
title: "AntiBiotics for Children with Diarrha (ABCD) Simulation Data Exploration"
author: "Allison Codi"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
subtitle: DATA 550
---

```{r, here-i-am, include=FALSE}
here::i_am(
  "report.Rmd"
)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

## Background
Acute watery diarrheal disease is a major contributor to child mortality, accounting for approximately 450,000 deaths in children under age 5 each year.
Severe diarrheal episodes are also associated with linear growth faltering which can pose risks of chronic disease, lower attained schooling, and limited economic success that last into adulthood.
Children who are dehydrated and/or malnourished in low-resource settings are at an elevated risk for mortality and linear growth faltering. 

Evidence has suggested that a large proportion of episodes in this population have a have a bacterial cause, thus could be treated with an antibiotic. 
However, current World Health Organization (WHO) guidelines do not broadly recommend treatment of acute watery diarrhea with antibiotics. 
AntiBiotics for Children with Diarrhea (ABCD) was a large, multi-country randomized control trial for azithromycin (an antibiotic) in dehydrated and/or malnourished children ages 2-23 months that aimed to assess whether the addition of azithromycin to diarrheal case management protocol in this population could reduce mortality and improve linear growth.
The initial trial was stopped early for futility, but detected a small reduction in linear growth faltering in the azithromycin (treatment) group.
A secondary analysis demonstrated larger azithromycin treatment effects among diarrheal episodes that were attributed to bacterial causes by qPCR quantity cut-offs.
Currently, my work aims to use causal inference and machine learning methods to estimate Optimal Treatment Rules (OTRs) which will recommend treatment to patients with a clinically relevant individual-level treatment effect. 
The goal is to create treatment guidelines that lead to better outcomes than current WHO recommendations, ultimately reducing mortality and improving linear growth. 

## Objectives 

For the scope of this assignment, I have simulated a dataset that mimics the real ABCD data.
I will describe the sociodemographic, illness, and malnutrition characteristics of the patients simulated in the dataset using appropriate tables. 
I will use histograms to describe the distribution of the pathogen quantities simulated in the dataset. 
Lastly, I will fit a linear regression model for length-for-age Z-score at day 90 of the trial (lazd90) based on clinically relevant select covariates in the dataset.
This will use the same model as I generated the dataset with, so we will expect the predicted values to be very close to the true values. 

## Dataset

The simulated dataset contains 6692 observations with 1 treatment assignment, 28 covariates, and 1 outcome variable. 
The treatment assignment is a binary variable corresponding to 0 if the child received placebo or 1 if the child received azithromycin. 
The outcome variables is length-for-age Z-score 90 days after enrollment.
Length-for-age Z-score can be used to identify linear growth faltering.
There is missingness at random present in lazd90 in align with the true ABCD dataset. 
The covariates in the simulated dataset include pathogen quantities, illness characteristics, sociodemographic characteristics, and malnutrition indicators. 
All covariates included in the dataset are listed in Tables 1-3.

```{r, load-abcd-data, echo = FALSE}
# Load dataset (used to grab labels for inline r code later on)
source("R/00_load_abcd_data.R")
```

# Descriptive Statistics for ABCD Simulation Dataset

## Sociodemographic Characteristics

```{r, socio-demographic-table, echo=FALSE}

readRDS(
  file= here::here("output/table_1.Rds")
)

```


Table 1 describes sociodemographic characteristics of the simulated ABCD study population for participants treated with azithromycin, participants who received placebo, and overall.
The categorical sociodemographic characteristics include sex, socioeconomic quintile, study site, enrollment month, and an indicator for enrollment during rotavirus season. 
Categorical covariates were described using N participants and the percentage of participants in each subgroup (within the given treatment strata).
The continuous covariates include age in months and number of children in the patient's household under age 5 years.
Continuous covariates were described using mean, standard deviation, median, min, and max. 

Note the study was designed as a Randomized Control Trial and data were simulated to mimic this.
Thus, sociodemographic characteristics are distributed evenly between strata. 

## Illness Characteristics

```{r, illness-characteristics, echo=FALSE}
readRDS(
  here::here("output/table_2.Rds")
)
```


Table 2 describes illness characteristics of the simulated ABCD study population for participants treated with azithromycin, participants who received placebo, and overall.
The categorical illness characteristics include an indicator for if the child has vomited from their diarrheal episode and level of dehydration at enrollment.
Categorical covariates were described using N participants and the percentage of participants in each subgroup (within the given treatment strata).
The continuous covariates include number of loose and solid stools in the 24 hour period prior to enrollment.
Continuous covariates were described using mean, standard deviation, min, and max. 

Note the study was designed as a Randomized Control Trial and data were simulated to mimic this.
Thus, illness characteristics are distributed evenly between strata. 

## Malnutrition Characteristics

```{r, malnutrition-characteristics, echo=FALSE}
readRDS(
  here::here("output/table_3.Rds")
)
```


Table 3 describes malnutrition characteristics of the simulated ABCD study population for participants treated with azithromycin, participants who received placebo, and overall.
Each malnutrition characteristic is continuous.
They include average middle-upper arm circumference, length for age z-score, weight for age z-score, and weight for length z-score at enrollment. 
The covariates were described using mean, standard deviation, median, min, and max. 

Note the study was designed as a Randomized Control Trial and data were simulated to mimic this.
Thus, illness characteristics are distributed evenly between strata. 

# Single Pathogen Quantity Visualizations

```{r, rotavirus, echo=FALSE, message=FALSE}
knitr::include_graphics(
  here::here("output/figure_1.png")
)
```

```{r, norovirus, echo=FALSE, message=FALSE}
knitr::include_graphics(
  here::here("output/figure_2.png")
)
```

```{r, adenovirus, echo=FALSE, message=FALSE}
knitr::include_graphics(
  here::here("output/figure_3.png")
)
```

```{r, astrovirus, echo=FALSE, message=FALSE}
knitr::include_graphics(
  here::here("output/figure_4.png")
)
```

```{r, sapovirus, echo=FALSE, message=FALSE}
knitr::include_graphics(
  here::here("output/figure_5.png")
)
```

```{r, st_etec, echo=FALSE, message=FALSE}
knitr::include_graphics(
  here::here("output/figure_6.png")
)
```

```{r, shigella, echo=FALSE, message=FALSE}
knitr::include_graphics(
  here::here("output/figure_7.png")
)
```

```{r, campylobacter, echo=FALSE, message=FALSE}
knitr::include_graphics(
  here::here("output/figure_8.png")
)
```


```{r, tepec, echo=FALSE, message=FALSE}
knitr::include_graphics(
  here::here("output/figure_9.png")
)
```


```{r, v_cholerae, echo=FALSE, message=FALSE}
knitr::include_graphics(
  here::here("output/figure_10.png")
)
```


```{r, salmonella, echo=FALSE, message=FALSE}
knitr::include_graphics(
  here::here("output/figure_11.png")
)
```

```{r, crypto, echo=FALSE, message=FALSE}
knitr::include_graphics(
  here::here("output/figure_12.png")
)
```

Figures 1-12 allow us to visualize the distribution of pathogen quantity for each pathogen measured in the ABCD trial.
The x-axes represent the transformed CT quantity of the pathogen detected in the sample.
The y-axes represent the number of simulated participants with a pathogen quantity in each bin.

The results are zero-inflated as most children did not have a given pathogen detected. 

```{r, most-freq-least-freq, echo=FALSE, message=FALSE}

summary_stats <- read.csv("output/summary_stats.csv")

which_path_max <- summary_stats$covariate[summary_stats$num_gt_0 == max(summary_stats$num_gt_0)]
which_path_max <- summary_stats$label[summary_stats$covariate == which_path_max]
which_path_max_val <- max(summary_stats$num_gt_0)

which_path_min <- summary_stats$covariate[summary_stats$num_gt_0 == min(summary_stats$num_gt_0)]
which_path_min <- summary_stats$label[summary_stats$covariate == which_path_min]
which_path_min_val <- min(summary_stats$num_gt_0)

```

`r  which_path_max` was detected most frequently with `r which_path_max_val` children infected.
`r  which_path_min` was detected least frequently with `r which_path_min_val` children infected.

# Model Fitting

The data for length for age z-score 90 days after enrollment was generated using the following model:

\begin{equation}
\begin{aligned}
lazd90 = \beta_0 + \beta_1dy1scrndiardays + \beta_2site + \beta_3ansesquintile + \beta_4agemchild + \\ \beta_5lfazscore + \beta_6I(shigella > 0) + \beta_7angrp01 + \beta_8I(shigella > 0)\times angrp01 + \\ \beta_9I(shigella > 0)\times lfazscore \times angrp01 + \beta_{10}I(shigella > 0) \times lfazscore \times angrp01 \times agemchild + \epsilon
\end{aligned}
\end{equation}
                                   
Where `dy1_scrn_diardays` is `r attr(abcd_data$dy1_scrn_diardays, "label")`, 
`site` is `r attr(abcd_data$site, "label")`, 
`an_ses_quintile` is `r attr(abcd_data$an_ses_quintile, "label")`,
`agemchild` is `r attr(abcd_data$agemchild, "label")`,
`lfazscore` is `r attr(abcd_data$lfazscore, "label")`,
`I(shigella > 0)` is an indicator that `r attr(abcd_data$shigella_new, "label")` is detected in the sample, 
and `an_grp_01` is a binary treatment indicator for Azithromycin (`an_grp_01 = 0` if Placebo, `an_grp_01 = 1` if Azithromycin).

I fit a this model using glm and plot the true simulated values for length for age z-score compared to the predicted values.

```{r, glm-fig, echo=FALSE, message=FALSE, warning=FALSE}
knitr::include_graphics(
  here::here("output/figure_13.png")
)
```

Figure 13 plots the true values for length-for-age Z-score 90 days post enrollment in the ABCD simulation dataset on the X-axis, and the predicted values for the length-for-age Z-score 90 days post enrollment in the ABCD simulation dataset on the Y-axis.
The Actual vs Predicted plot of length-for-age Z-score 90 days post enrollment shows our model is a good fit for the dataset (as expected given it is the same model we generated the data from).



